---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# (╯°□°）╯︵ ┻━┻ {flippingtables} (╯°□°）╯︵ ┻━┻

> All data.frame print methods are wrong, but some are useful

<!-- badges: start -->
<!-- badges: end -->

Inside you are two R users:

  - One seeks meaning and truth in the numbers
  - One seeks correctness and reproducibility in the code

The problem with all `print()` methods is that none can satisfy both users.

Methods like `paint::paint()`, `pillar::glimpse()`, and `str()` are pitched at
the developer. They aim to succinctly display types, and convey a feel for the data with examples.

The default `print.data.frame()` aims to give the analyst all the numbers, and even does an okay job
of it, provided you have a ultra-wide screen monitor, and less than a couple of hundred rows.

Others try to compromise between the two modalities, and they are generally the
worst to use. They'll consider the volume of data and space available, and
concoct methods of truncation. Not rows, nor columns, nor column names, nor
decimal places are held sacred. Using these involves a lot of retracing your steps
fiddling with print arguments or setting options. It's enough to make you want
to flip tables.

You need multiple ways of printing data to suit whatever kind of user you happen
to be today, and you need them arrayed easily at your fingertips to keep you in
the zone. This package places the way you view your data in your session fully
within your control.

## Installation

``` r
remotes::install_github("MilesMcBain/flippingtables")
```

## How to flip() tables

The setup is:

1. Create a configuration (probably in .Rprofile) with `register_flips()`

2. Enable the config with `flip_on()` (possibly in .Rprofile)
  - may seem redundant but thank me when you need to `flip_off()` custom print methods because they balk at exotic text
   encodings or some such.

3. Use `flip()` to cycle between configured print methods for configured classes.
  - Probably via a keyboard shortcut

## Configuration

Here's an example config:

```{r configuration}
library(flippingtables)
register_flips(
  printer_fns = list(
    paint::paint, # a pretty good option if I do say so myself.
    function(x) default_print(x, n = 100), # a long format
    function(x) withr::with_options(list(width = 300), default_print(x, n = 100))
  ), # a wide format
  printed_classes = list(
    print_override(class = "tbl", pkg_namespace = "pillar"),
    print_override(class = "data.frame", pkg_namespace = "base"),
    print_override(class = "data.table", pkg_namespace = "data.table")
  )
)
flip_on() # now it's live!
```

First in `printer_fns` we declare the print methods we want be able to toggle
between for use as the automatic `print()` in our R consolse sessions. Using
anonymous functions is a great way to prototype custom print methods. The
special function `print_default()` stands in for whatever the default print for
the class being printed would normally be.

Then we nominate `tbl`, `data.frame`, and `data.table` as 'flippable'
classes. We have to nominate the 'top-level' class that has the print method
we want to override. I.e. even though `tbl` (a generic `tibble` used by
`{pillar}`) and `data.table` are also `data.frame` we can't just configure
`data.frame` because objects of those classes have the `"tbl"` or `"data.table"`
appear earlier in their vector of classes. This means their corresponding `print()`
methods dispatched instead of `print.data.frame` (or the custom print method we routed that too).

### Flipping

Calling `flippingtables::flip()` will advance the binding for the current print method to be
used for all configured classes to the next one in the list. Print methods are cycled through in the order
they are configured, and of course the cycle wraps-around so cycling can happen
endlessly.

If the last result (`.Last.value`) has a class that is configured for flipping,
then the object is automatically reprinted with the new print method selected
the object is automatically reprinted with the new print method selected by
`flip()`.

Here's how it might look:

```{r example, eval=FALSE}
# assuming config above
library(palmerpenguins)
penguins

# tibble [344, 8]
# species           fct Adelie Adelie Adelie Adelie Adelie Ad~
# island            fct Torgersen Torgersen Torgersen Torgers~
# bill_length_mm    dbl 39.1 39.5 40.3 NA 36.7 39.3
# bill_depth_mm     dbl 18.7 17.4 18 NA 19.3 20.6
# flipper_length_mm int 181 186 195 NA 193 190
# body_mass_g       int 3750 3800 3250 NA 3450 3650
# sex               fct male female female NA female male
# year              int 2007 2007 2007 2007 2007 2007

flip()

#  A tibble: 344 × 8
#     species island    bill_length_mm bill_depth_mm flipper_length_mm body_mass_g
#     <fct>   <fct>              <dbl>         <dbl>             <int>       <int>
#   1 Adelie  Torgersen           39.1          18.7               181        3750
#   2 Adelie  Torgersen           39.5          17.4               186        3800
#   3 Adelie  Torgersen           40.3          18                 195        3250
#   4 Adelie  Torgersen           NA            NA                  NA          NA
#   5 Adelie  Torgersen           36.7          19.3               193        3450

#   .. Output continues ..

#  95 Adelie  Dream               36.2          17.3               187        3300
#  96 Adelie  Dream               40.8          18.9               208        4300
#  97 Adelie  Dream               38.1          18.6               190        3700
#  98 Adelie  Dream               40.3          18.5               196        4350
#  99 Adelie  Dream               33.1          16.1               178        2900
# 100 Adelie  Dream               43.2          18.5               192        4100
# # ℹ 244 more rows
# # ℹ 2 more variables: sex <fct>, year <int>
# # ℹ Use `print(n = ...)` to see more rows

# it's sticky!
penguins

#  A tibble: 344 × 8
#     species island    bill_length_mm bill_depth_mm flipper_length_mm body_mass_g
#     <fct>   <fct>              <dbl>         <dbl>             <int>       <int>
#   1 Adelie  Torgersen           39.1          18.7               181        3750
#   2 Adelie  Torgersen           39.5          17.4               186        3800
#   3 Adelie  Torgersen           40.3          18                 195        3250
#   4 Adelie  Torgersen           NA            NA                  NA          NA
#   5 Adelie  Torgersen           36.7          19.3               193        3450

#   .. Output continues ..

#  95 Adelie  Dream               36.2          17.3               187        3300
#  96 Adelie  Dream               40.8          18.9               208        4300
#  97 Adelie  Dream               38.1          18.6               190        3700
#  98 Adelie  Dream               40.3          18.5               196        4350
#  99 Adelie  Dream               33.1          16.1               178        2900
# 100 Adelie  Dream               43.2          18.5               192        4100
# # ℹ 244 more rows
# # ℹ 2 more variables: sex <fct>, year <int>
# # ℹ Use `print(n = ...)` to see more rows

flip()

# # A tibble: 344 × 8
#     species island    bill_length_mm bill_depth_mm flipper_length_mm body_mass_g sex     year
#     <fct>   <fct>              <dbl>         <dbl>             <int>       <int> <fct>  <int>
#   1 Adelie  Torgersen           39.1          18.7               181        3750 male    2007
#   2 Adelie  Torgersen           39.5          17.4               186        3800 female  2007
#   3 Adelie  Torgersen           40.3          18                 195        3250 female  2007
#   4 Adelie  Torgersen           NA            NA                  NA          NA NA      2007
#   5 Adelie  Torgersen           36.7          19.3               193        3450 female  2007
#
# .. Output continues ..
#
#  95 Adelie  Dream               36.2          17.3               187        3300 female  2008
#  96 Adelie  Dream               40.8          18.9               208        4300 male    2008
#  97 Adelie  Dream               38.1          18.6               190        3700 female  2008
#  98 Adelie  Dream               40.3          18.5               196        4350 male    2008
#  99 Adelie  Dream               33.1          16.1               178        2900 female  2008
# 100 Adelie  Dream               43.2          18.5               192        4100 male    2008
# # ℹ 244 more rows
# # ℹ Use `print(n = ...)` to see more rows
```

### Keyboard shortcut

In VSCode:



``` json
{
      "description": "flip() between print methods",
      "key": "<CHOOSE A KEYBINDING>",
      "command": "r.runCommand",
      "when": "editorTextFocus",
      "args": "flippingtables::flip()"
}
```

In RStudio:

  - `flip()` is exposed as an RStudio addin, which can be bound to a keyboard shortcut. See [instructions](https://docs.posit.co/ide/user/ide/guide/productivity/custom-shortcuts.html)

In other editor:

  - I think you should mostly be fine since it doesn't depend on any editor state, but let me know if I can expose anything that makes it easier.

### Cookbook

On of my personal favourites for looking at sumarised data analytically
(provided it is sorted sensibly) is `knitr::kable()`. It has the added benefit
that you can easily copy-paste the output to make nice tables in applications
that speak markdown. It presents two problems for use directly:

- it doesn't return the data it printed invisibly, unlike most methods intended to be used
for this purpose. This will cause printing the `.Last.value` with flip() to stop as soon as this
method is hit.
- it returns an object of a different class which has its own print method. R
console seems not to like the idea of running through the print dispatch again,
so we get nothing.

We work around this like so:

```{r, eval=FALSE}
library(flippingtables)
register_flips(
  printer_fns = list(
    paint::paint, # a pretty good option if I do say so myself.
    function(x) {
      print(knitr::kable(x))
      invisible(x)
    }
  ),
  printed_classes = list(
    print_override(class = "tbl", pkg_namespace = "pillar"),
    print_override(class = "data.frame", pkg_namespace = "base"),
    print_override(class = "data.table", pkg_namespace = "data.table")
  )
)
flip_on() # now it's live!
flip()
penguins

# |species   |island    | bill_length_mm| bill_depth_mm| flipper_length_mm| body_mass_g|sex    | year|
# |:---------|:---------|--------------:|-------------:|-----------------:|-----------:|:------|----:|
# |Adelie    |Torgersen |           39.1|          18.7|               181|        3750|male   | 2007|
# |Adelie    |Torgersen |           39.5|          17.4|               186|        3800|female | 2007|
# |Adelie    |Torgersen |           40.3|          18.0|               195|        3250|female | 2007|
# .. Output continues ..
```
